---
title: "Class 6: R Functions"
author: "Reta Ado (PID: A17814740)"
format: pdf  
toc: true 
---

## Background

All functions in R have at least 3 things:

-   a *name* (we pick that and use it to call the function)
-   input *arguments* (one or more comma separaed inputs that go inside the brackets when we call the function),
-   the "body" (the lines of R code that do the work of the function).


> Q1a: Your first version of the function should add two input numbers together. For example, add(4,
7) should return 11. 

## A first function

Here we will create a function to add some numbers. Let's call it `add()`

Input arguments can be either **required** or **optional**. The later have fall-back default that will be used if the user does not specify them.

```{r}
add <- function(x, y = 0) { 
  x + y 
}
```

Can we use our new function:

```{r}
add(10, 100) 
```

```{r}
add(10) 
```

> Q1b: For you second version, adapt your first function so it can take a single input vector or two inputs as before. For example, add(4, 7) and add( c(4,7,10) ).


```{r}
add <- function(x, y = 0) { 
 sum(x, y) 
} 
```

```{r}

add(4, 7) 
add(c(4,7, 10), ) 
```

> Q1c: Finally, on your own (outside of class) create a third version of your function that can add any number of inputs that the user provides. For example, add(1, 2, 3, -4) should return 2. Hint: explore the ... (dots) argument or the base R function sum(). 

```{r}
add <- function(...) {
  sum(...)
}

```

```{r}
add(1, 2, 3, -4)
```

We can explicitly set a **return** value output from a function (rather than the default last line) by using the `return()` function call.

```{r}
add<- function(x, y=0, z=0) {
  sum(x, y, z) 
  cat("Is it break time yet\n")
  } 

add(10,100) 
```

## A generate_data() function

A useful function here is the "base R" `sample()` function:

```{r}
sample(1:5, size=3) 
```

We can use this to make a random nucelotide seuence if we draw from "A", "C", "G", and "T"...

```{r}

sample(1:5, size=60, replace = TRUE) 

```

```{r}
sample(x=c("A", "C", "G", "T"), size=10, replace = TRUE) 
```

> Q2a: Write a function `generate_dna()` that returns a random DNA sequence of a length specified by the user. Your first version should return a multi-element vector of single character nucleotides. For example generate_dna(6) might return "A", "T", "T", "G", "A", "C". 

```{r}
generate_dna <- function(len) {
  sample(x=c("A", "C", "G", "T"), size=len, replace = TRUE) 
} 
```

```{r}
generate_dna(len=100) 
generate_dna() 

```

> Q2b: Your second version should optionally be able to return either a multi-element vector of single character nucleotides (as before) or a single character string (not a vector of single letters but a singe vector of multiple letters). For example "AAGGCTG". 

```{r}
generate_dna <- function(len, single.element = FALSE) {
  ans <- sample(x = c("A", "C", "G", "T"), size = len, replace = TRUE)
  
  if (single.element == TRUE) {
    ans <- paste(ans, collapse = "")
  }
  
  return(ans)
}

 
```

```{r}
#example of expected behavior 
generate_dna(6)

#example of single character string 
generate_dna(6, single.element = TRUE)
```

Functions that could be useful here are `paste()`. `if(),`cat(), and `return()`

```{r}
paste(c("A", "C", "G"), collapse = "---" ) 
```

> Q2c: Finally, create a final version of your function that prints out a FASTA format sequence with an id line indicating the sequence length.

```         
len9
CGAAGGCTG
```

```{r}
cat("hello \t there") 
```

```{r}
generate_dna <- function(len=10, single.element=TRUE) {
  
  ans<- sample(x=c("A", "C", "G", "T"), size=len, replace = TRUE)
  #cat("Hello....")
  
  if(single.element) {
    #cat("is it me you are looking for...")
    ans <- paste( ans, collapse = "")
  }
  
  ## Format as FASTA with an ID line
  
  cat( paste(">len", len, "\n", sep=""))
  cat(ans)
  cat("\n")
  ##
  
  return(ans)
}
```

```{r}
x <- generate_dna(20)
```

## Write a \`generate_protein() function

> Q3: Write a function generate_protein() that returns a random peptide/protein sequence of a length specified by the user. For example generate_protein(6) might return "WQRTAG".

```{r}
generate_protein <- function(len) {
  
  # Write out all 20 standard amino acids using single letter code
  aa <- c("A", "R", "N", "D", "C", "E", "Q", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V") 
  
  # Randomly  sample amino acids with replacement to build a protein sequence
  # size = len means the sequence will be as long as the user specifies
  ans <- sample(x = aa, size = len, replace = TRUE)
  
  # Collapse the vector into a single string 
  paste(ans, collapse = "")
 
} 

```

```{r}
# Example of the expected behavior
generate_protein(6) # e.g. "WQRTAG" 
```

> Q4: Adapt and use your generate_protein() function to generate a series of random protein sequences ranging from 6 to 13 amino acids in length (one sequence per length). Take advantage of the base R function for() or sapply() so that you do not have to call generate_protein() eight times by hand.

```{r}
for(l in 6:13) { 
  cat(">id.", l, "\n", sep = "") 
  cat(generate_protein(l), "\n")  
  
  
} 
```

```{r}
generate_protein(6) 
```

# Are Our peptides "unique in nature?"

> Q5: Take your FASTA-formatted peptides from Q4 and run them as a single BLASTp search against the Non-redundant protein sequences (nr) database at https://blast.ncbi.nlm.nih.gov/. For this question do not restrict the organism (leave the Organism field blank so that the full nr database is searched).


| Length  | Ide | Cov   | Uniq |
|---------|------|------|------|
|  6      | 100  | 100  |  N   |
|  7      | 100  | 100  |  N   |
|  8      | 100  | 100  |  N   |
|  9      | 100  | 89   |  Y   |
|  10     |88.89 | 90   |  Y   |
|  11     |81.82 | 100  |  Y   |
|  12     | 100  | 75   |  Y   |
|  13     |81.82 | 85   |  Y   |

>Q5a: At which sequence length do your randomly generated peptides start to look “unique in nature” (i.e. no 100% coverage + 100% identity hit)? 

They start to look unique in nature after the 8th sequence length (starts at 9). 

>Q5b: Speculate why very short random peptides are almost always found in nr, while longer ones typically are not. Your answer should refer both to the size of the sequence space (20𝐿 for a peptide of length 𝐿) and to the size of the known protein universe. 

Short random peptides are almost always found in nr  because their sequence space (20^L) is very small compared to *n* (number of sequences in known protein universe), which makes it likely that any randomly generated sequence has been sampled by a natural protein. As peptide length (L) increases, the sequence space grows exponentially while the number of known proteins *n* remains fixed. Once sequence space exceeds *n*, the vast majority of possible sequences haven't appeared in any natural protein, making it increasingly unlikely that a randomly generated peptide of length L would be found in the database. 


> Q6: In 3–6 sentences total and using your Q5 data and the reasoning from Q5b, what do you think this minimum length is and why might it be a bad design choice for the immune system to present very short peptides?  


The minimum useful peptide length is around 9 amino acids since this is where sequences first become unique in nature. Presenting very short peptides (like 6,7,8 in my example) would be a bad design choice for the immune system because their sequence space is too small, meaning those sequences are too short to be distinctive from naturally occurring proteins. Therefore, the body's immune system does not have ability to determine pathogens from self peptides, triggering possibly bad autoimmune responses. Longer peptides have a  larger sequence space (ex:20^9), making it much more likely that a peptide is unique to a pathogen rather than shared with a host protein.   



