---
title: "Class 7: Machine Learning 1"
author: "Reta Ado (PID: A17814740)" 
date: 21 April 2026 
format: pdf  
toc: true 
---

## Background 

Today we will explore some core machine learning methods that are very popular in bioinformatics. These include **clustering** and **dimensionallity reduction**. 


## K-means clustering 

The main function in "base" R for K-means clustering is called `kmeans()` 

Before we go to deep let's makeup some "simple" data that we can cluster and know if we are getting a good answer or not. To do this we can use the `rnorm()` function: 

```{r}
hist(rnorm(10000, mean = 3))  
```


```{r}
x <- c(rnorm(30, -3), rnorm(30, +3)) 

z <- cbind(x=x, y=rev(x)) 
plot(z) 
```

```{r}
z <- cbind(x=x, y=rev(x)) 
z 
```


```{r}
p <- 1:5 

rev(p) 

```

Now we can run `kmeans()` on this input `z` and see what the results look like. 

```{r}
km <- kmeans(z, centers = 2) 
km 
```

```{r}
attributes(km) 
```

> Q. How many points are in each cluster? 

```{r}
km$size 
```


> Q: What component of your result object details cluster assignment/membership?

```{r}
km$cluster 
```


> Q. What component of your result object details cluster center? 


```{r}
km$centers 
```


> Q. Plot `z` colored by the kmeans cluster assignment and add cluster centers as blue points 

```{r}
plot(z, col = km$cluster)
points(km$centers, col = "blue", pch = 15) 
```


> Q. Run a K-means clustering and plot the results asking for 4 clusters (K=4)? 

```{r}
km4 <- kmeans(z, centers = 4) 
plot(z, col = km4$cluster)
points(km4$centers, col = "black", pch = 15) 
```

> **N.B.** You need to tell K-means the number of clusters (i.e. set `centers = 2`)!! 


One approach is to try different values for `centers` and then pick the best... 

```{r}
ans <- NULL 

for(i in 1:10) { 
  km <- kmeans(z, centers = i)
  ans <- c(ans, km$tot.withinss)  
} 

plot(ans, typ = "o", xlab = "Number of clusters", ylab = "Total Sum of Squars Distance") 

```


## Hierarchial Clustering 

The main function in "base" R for Hierarchial Clustering is called `hclust()` 

This function does not take your "raw" data for clustering. You must first build a "distance matrix" from your data and pass this as input to `hclust()` 

```{r}

d <- dist(z) 
hc <- hclust(d) 
hc 

```

There is a bespke `plot()` method for `hclust()` result objects. 

```{r} 

plot(hc) 
abline(h=8, col = "red") 

```


Once we have our `hclust` object (our "tree" of "cluster dendrogram") we can *"cut"* the tree to reveal the clustering pattern. 

```{r}
cutree(hc, k =4) 

```


> Q. Make a plot of `z` with your hclust results (i.e. colored by cluster membership) 

```{r}
grps <- cutree(hc, k = 2) 
plot(z, col = grps) 
```


## Principal Component Analysis (PCA) 

PCA is a dimensionallity reduction method that is popular for revealing patterns in complex datasets 

## Analysis of UK food data 

Let's look at some data on the eating habits of folks from the UK to see if there are patterns and trends that have sone countries regions being distinct from others. 

## Data Input 

The data is made available in CSV format so we can use the `read.csv(url)` 
```{r}
# Store the URL pointing to the UK foods dataset (hosted as a CSV file)
url <- "https://tinyurl.com/UK-foods"
# Read CSV 
x <- read.csv(url)  
x 
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

The data frame x has 17 rows and 5 columns. 



```{r}
dim(x) 
```

```{r}
ncol(x) 
nrow(x)
```


> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions? 

The R function that you can use  is dim(x) since it gives you both dimensions in one calk.There are  17 rows and 5 columns. There are technically 4 columns, but the data is not tidied yet to reflect the 4 countries as it considers x. 

## Tidy the data 

Fix anything that went wrong with data import. 

Option 1: 

```{r}
#Tidy data using option #1 
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```

Option 2: 

```{r}
#Tidy data using option #2 
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names = 1)   
x 
```

```{r}
dim(x) 
```


> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances? 

I prefer the second approach (row.names = 1) to the row-names problem as it is simpler. This is also more robust because it fixes the problem at the import. The option that removes a row is risky because if you run your code twice, it can keep removing rows. 

## Exporatory analysis 

Make some plots to help make sense of obvious trends... 

```{r}

#Plot a barplot of the data 
barplot(as.matrix(x), beside = TRUE, col = rainbow(nrow(x))) 

```

> Q3: Changing what optional argument in the above barplot() function results in the following plot? 

Changing beside=TRUE to beside=FALSE (or removing it) produces the barplot shown. 

```{r}
barplot(as.matrix(x), beside = FALSE, col = rainbow(nrow(x))) 
```

```{r}
head(x) 
```

```{r}

library(tidyr)

# Convert data to long format for ggplot with `pivot_longer()`
x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")

dim(x_long)
```

```{r}
# Create grouped bar plot
library(ggplot2) 
```


```{r}
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_bw()
```


> Q4: Changing what optional argument in the above ggplot() code results in a stacked barplot figure? 

Changing position = "dodge" to position = "stack" in geom_col() produces the stacked barplot. 

```{r}
# Create stacked bar plot
library(ggplot2)

ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "stack") +   # "stack" accumulates bars vertically
  theme_bw() 
```


> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot? 

The pairs() function creates a matrix of scatter plots comparing food consumption between every pair of countries, with each colored dot representing one food category. When a point lies on the diagonal, it means both countries consume equal amounts of that food. Points above the diagonal mean the y-axis country consumes more, and points below mean the x-axis country consumes more. Overall, England, Wales, and Scotland show similar consumption patterns, while Northern Ireland deviates more. 

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16) 
```


```{r}
# Install pheatmap

library(pheatmap)

# Make a heatmap 
pheatmap( as.matrix(x) )
```

> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set? 

Based on the figures, England and Wales cluster most closely together, indicating very similar food consumption patterns, while Scotland sits nearby and Norther Ireland is the outlier. N. Ireland shows higher consumption of Fresh potatoes and Other meat, and lower consumption of Fresh fruit compared to the other three countries. However, the heatmap alone makes it somewhat difficult to identify all differences clearly, since the color scale is dominated by high-consumption foods like Fresh potatoes, which can obscure subtler variations in other categories. 

**Key-point**: Even relatively small datasets can prove challenging to interpret. 

## PCA 

The main function in "base" R for PCA is called `prcomp()`. This function wants the "observations" to be rows and the "variables" to be columns. 

So here we need to take the transpose of our `x` input object. 

```{r}

# Use the prcomp() PCA function 
pca <- prcomp(t(x))  
summary(pca) 
```


The returned `pca` object has components that we can use to make our main result figures: 



```{r}
# Find attributes 
attributes(pca) 
```
The main result figure from this analysis is called a "PC score plot" or PC1 vs PC2 plot." 

This plot shows how samples (in this case countries) relates to each oter along our new PC axis. 

Thisis our new "reduced-dimensional space." In this case 2 dimensions, PC1 and PC2, that capture most of the variance in the original 17 dimension data set. 


```{r}
library(ggplot2) 

ggplot(pca$x) + 
  aes(PC1, PC2) + 
  geom_point() 
```

```{r} 
mycols <- c("orange", "red", "blue", "darkgreen")

ggplot(pca$x) + 
  aes(PC1, PC2) + 
  geom_point(col=mycols) 
```


> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points. 

```{r}
# Answering Q7 

# Create a data frame for plotting
df <- as.data.frame(pca$x)
df$Country <- rownames(df)

# Plot PC1 vs PC2 with ggplot
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.5) +
  xlim(-270, 500) +
  xlab("PC1") +
  ylab("PC2") +
  theme_bw()

```


> Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document. 

```{r}

# Answering Q8 


# Plot PC1 vs PC2, mapping country names to both label and color aesthetics
ggplot(pca$x) +
  aes(x = PC1, y = PC2, label = rownames(pca$x), color = rownames(pca$x)) +
  geom_point(size = 3) +                       
  geom_text(vjust = -0.5) +                    
  scale_color_manual(values = c("England" = "orange",    
                                "Wales" = "red",         
                                "Scotland" = "blue", 
                                "N.Ireland" = "darkgreen")) +
  xlim(-270, 500) +                            
  xlab("PC1") +                                 
  ylab("PC2") +                                
  theme_bw()                                    
```

## Variable Loadings 

```{r}
pca$rotation 
```

```{r}

## Lets focus on PC1 as it accounts for > 90% of variance 
ggplot(pca$rotation) +
  aes(x = PC1, 
      y = reorder(rownames(pca$rotation), PC1)) +
  geom_col(fill = "steelblue") +
  xlab("PC1 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```


> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about? 

The two food groups that feature most prominently in PC2 are Soft_drinks ( positive loading) and Fresh_potatoes (negative loading ). PC2 mainly tells us about the contrast between these two foods, as countries that score high on PC2 consume more soft drinks, while countries that score low on PC2 consume more fresh potatoes. This explains why Scotland sits high on PC2 (higher soft drink consumption) while Wales sits low on PC2 (higher fresh potato consumption). 

```{r}
# Generate loadings plot for PC2
ggplot(pca$rotation) +
  aes(x = PC2,
      y = reorder(rownames(pca$rotation), PC2)) +
  geom_col(fill = "steelblue") +
  xlab("PC2 Loading Score") +
  ylab("") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 9))
```
















